---
title: "Reconstructing the English Indices of Deprivation 2025 with COINr"
description: "A reproducible LSOA-level walkthrough of the IMD 2025 transformations, hierarchy and aggregation"
author: "IMD package contributors"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-copy: true
    df-print: kable
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
vignette: >
  %\VignetteIndexEntry{Reconstructing the English Indices of Deprivation 2025 with COINr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Aim and scope

The English Index of Multiple Deprivation 2025 (IMD 2025) combines seven domains of deprivation for 33,755 Lower-layer Super Output Areas (LSOAs) in England. This walkthrough reconstructs the index with [{COINr}](https://bluefoxr.github.io/COINr/), an R package for building and analysing composite indicators.

The official method has three broad stages:

1. combine indicators into domain or sub-domain scores;
2. rank each domain and map those ranks to a common exponential distribution; and
3. combine the seven transformed domains using the official domain weights.

The [IMD 2025 Technical Report](https://assets.publishing.service.gov.uk/media/68ff59c80f801e57b5bef907/ID_2025_Technical_Report.pdf) describes the method in detail. The source workbooks are available from the [English indices of deprivation 2025 release page](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2025).

::: {.callout-important title="This is a public-data reconstruction, not a fully independent replication"}
A completely independent reconstruction is not possible from the released data. The Income and Employment domains require shrinkage inputs that are not all published. Four pupil-derived indicators in the Children and Young People education sub-domain are also withheld because they derive from pupil microdata.

We therefore use the published Income and Employment domain scores, and the published Children and Young People sub-domain score, at the points where the public data run out. Health, Crime, Barriers to Housing and Services, Living Environment, the final exponential transformation, and the final IMD aggregation are reconstructed from released inputs.
:::

## Packages and source files

We need only three R packages. `{COINr}` manages the composite-indicator objects, metadata, normalisation and aggregation. `{dplyr}` handles data manipulation, and `{readxl}` reads the official workbooks.

```{r}
#| label: setup

library(COINr)
library(dplyr)
library(readxl)
```

The reconstruction uses three official files:

- File 7, containing published scores, ranks, deciles and population denominators;
- File 8, containing the released underlying indicators; and
- File 9, containing exponentially transformed domain scores for final-stage validation.

The code below downloads a file only when it is not already present. Set the `IMD2025_DATA_DIR` environment variable to use a persistent local directory. Otherwise, the files are stored in a temporary directory for the current R session.

```{r}
#| label: source-files

data_dir <- Sys.getenv(
  "IMD2025_DATA_DIR",
  unset = file.path(tempdir(), "imd2025")
)
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

source_urls <- c(
  all_results = paste0(
    "https://assets.publishing.service.gov.uk/media/691ded56d140bbbaa59a2a7d/",
    "File_7_IoD2025_All_Ranks_Scores_Deciles_Population_Denominators.csv"
  ),
  indicators = paste0(
    "https://assets.publishing.service.gov.uk/media/691dec012c6b98ecdbc500d4/",
    "File_8_IoD2025_Underlying_Indicators.xlsx"
  ),
  transformed_domains = paste0(
    "https://assets.publishing.service.gov.uk/media/691ded670dcbf6343e9a2a6c/",
    "File_9_IoD2025_Transformed_Domain_Scores.xlsx"
  )
)

source_paths <- file.path(data_dir, basename(source_urls))
names(source_paths) <- names(source_urls)

for (source_name in names(source_urls)) {
  if (!file.exists(source_paths[[source_name]])) {
    download.file(
      source_urls[[source_name]],
      source_paths[[source_name]],
      mode = "wb",
      quiet = TRUE
    )
  }
}
```

::: {.callout-note title="For an offline package vignette"}
Package checks should not depend on live downloads. Before adding this file to `{IMD}`, either turn Files 7--9 into documented package data objects or place them under `inst/extdata/imd2025/`. The three paths above can then be replaced with `system.file()` calls. See [Package integration](#package-integration) at the end.
:::

## Translate the official transformations into R

### Rank-based normal scores

Most indicators are measured in different units. Before they are combined, the official method ranks each indicator and converts its rank to a standard normal score:

$$
z_i = \Phi^{-1}\left(\frac{r_i - 0.5}{N}\right),
$$

where $r_i$ is the indicator rank, $N$ is the number of observed LSOAs, and $\Phi^{-1}$ is the standard normal quantile function. Average ranks are used for ties.

```{r}
#| label: normal-rank-function

imd_rank_normal <- function(x) {
  out <- rep(NA_real_, length(x))
  observed <- !is.na(x)
  n <- sum(observed)

  out[observed] <- qnorm(
    (rank(x[observed], ties.method = "average") - 0.5) / n
  )

  out
}
```

This function satisfies the requirements for a custom `COINr::Normalise()` function: it takes a numeric vector and returns a numeric vector of the same length.

### Exponential transformation

Before domains or sub-domains are combined, the official method often maps their ranks to a common exponential distribution:

$$
X = -23\ln\left[1-R\left(1-e^{-100/23}\right)\right],
$$

where $R$ is the fractional rank, running from $1/N$ for the least deprived LSOA to 1 for the most deprived. The result ranges from approximately 0 to 100 and deliberately spreads out the deprived end of the distribution.

```{r}
#| label: exponential-function

imd_rank_exponential <- function(x, constant = 23) {
  out <- rep(NA_real_, length(x))
  observed <- !is.na(x)
  n <- sum(observed)

  rank_fraction <-
    rank(x[observed], ties.method = "average") / n

  out[observed] <- -constant *
    log(
      1 - rank_fraction * (1 - exp(-100 / constant))
    )

  out
}
```

### A weighted sum that preserves the published scale

The seven published domain weights sum to 0.999 rather than 1 because the three smallest weights are displayed as 9.3%. The released IMD score applies these displayed weights directly. `COINr::a_amean()` divides by the sum of the weights, so we use a small custom aggregation function to reproduce the published score scale exactly.

```{r}
#| label: weighted-sum-function

imd_weighted_sum <- function(x, w) {
  if (anyNA(x)) {
    return(NA_real_)
  }
  sum(x * w)
}
```

This distinction affects the score scale slightly but does not affect ranks.

## A small COINr constructor

A COINr `coin` needs two tables:

- `iData`, with one row per LSOA and one column per indicator; and
- `iMeta`, describing each indicator's name, direction, weight and parent aggregate.

The helper below builds a two-level coin, optionally normalises its indicators, aggregates them, and returns both the coin and its aggregate score. Building separate coins for different stages is useful here because IMD applies different transformations at different points in its hierarchy.

```{r}
#| label: coin-constructor

meta <- function(iCode, iName, Direction = 1, Weight = 1) {
  tibble(iCode, iName, Direction, Weight)
}

make_coin <- function(
  iData,
  indicator_meta,
  aggregate_code,
  aggregate_name,
  normaliser = NULL
) {
  stopifnot(
    identical(names(iData)[1], "uCode"),
    identical(names(iData)[-1], indicator_meta$iCode)
  )

  iMeta <- bind_rows(
    indicator_meta |>
      transmute(
        Level = 1L,
        iCode,
        iName,
        Parent = aggregate_code,
        Direction,
        Weight,
        Type = "Indicator"
      ),
    tibble(
      Level = 2L,
      iCode = aggregate_code,
      iName = aggregate_name,
      Parent = NA_character_,
      Direction = 1,
      Weight = 1,
      Type = "Aggregate"
    )
  )

  coin <- new_coin(
    iData = iData,
    iMeta = iMeta,
    level_names = c("Indicator", "Aggregate"),
    quietly = TRUE
  )

  dset <- "Raw"

  if (!is.null(normaliser)) {
    coin <- suppressMessages(Normalise(
      coin,
      dset = "Raw",
      global_specs = list(f_n = normaliser)
    ))
    dset <- "Normalised"
  }

  coin <- suppressMessages(Aggregate(
    coin,
    dset = dset,
    f_ag = "imd_weighted_sum"
  ))

  list(
    coin = coin,
    score = get_dset(coin, "Aggregated")[[aggregate_code]]
  )
}
```

COINr automatically handles direction. A direction of `1` means that higher values represent more deprivation; `-1` means the reverse.

## Read and align the official data

LSOA codes are the join key. We sort every input by code and stop if any sheet does not align with File 7. This avoids the silent row-order errors that are particularly dangerous when working with 33,755 observations.

```{r}
#| label: read-data

official <- read.csv(
  source_paths[["all_results"]],
  check.names = FALSE
) |>
  arrange(`LSOA code (2021)`)

uCode <- official$`LSOA code (2021)`
n_lsoa <- length(uCode)

read_indicator_sheet <- function(sheet) {
  read_excel(source_paths[["indicators"]], sheet = sheet) |>
    arrange(`LSOA code (2021)`)
}

assert_codes <- function(data, label) {
  if (!identical(data$`LSOA code (2021)`, uCode)) {
    stop(label, " LSOA codes do not align with File 7.")
  }
  invisible(data)
}

education <- assert_codes(
  read_indicator_sheet("IoD25 Education Domain"),
  "Education"
)
health <- assert_codes(
  read_indicator_sheet("IoD25 Health Domain"),
  "Health"
)
crime <- assert_codes(
  read_indicator_sheet("IoD25 Crime Domain"),
  "Crime"
)
barriers <- assert_codes(
  read_indicator_sheet("IoD25 Barriers Domain"),
  "Barriers"
)
living <- assert_codes(
  read_indicator_sheet("IoD25 Living Env Domain"),
  "Living Environment"
)

stopifnot(n_lsoa == 33755L)
```

## Reconstruct the domains

### Education, Skills and Training

The Education domain has two equally weighted sub-domains:

- Children and Young People; and
- Adult Skills.

The Children and Young People sub-domain uses five indicators with factor-analysis weights. Only Entry to Higher Education is released; the four pupil-derived indicators are not. We therefore start from the published Children and Young People sub-domain score in File 7. The Adult Skills indicator is available in File 8.

Both sub-domains are exponentially transformed before being combined with equal weights.

```{r}
#| label: education-domain

education_coin <- make_coin(
  data.frame(
    uCode,
    children_young_people = official$`Children and Young People Sub-domain Score`,
    adult_skills = education$`Adult skills and English language proficiency indicator`
  ),
  meta(
    c("children_young_people", "adult_skills"),
    c("Children and Young People", "Adult Skills"),
    Weight = c(0.5, 0.5)
  ),
  aggregate_code = "education",
  aggregate_name = "Education, Skills and Training",
  normaliser = "imd_rank_exponential"
)
```

### Health Deprivation and Disability

The four Health indicators are transformed to normal scores and combined using the factor-analysis weights reported in the Technical Report.

```{r}
#| label: health-domain

health_coin <- make_coin(
  data.frame(
    uCode,
    acute_morbidity = health$`Acute morbidity indicator`,
    illness_disability = health$`Comparative illness and disability ratio indicator`,
    mental_health = health$`Mental health indicator`,
    years_life_lost = health$`Years of potential life lost indicator`
  ),
  meta(
    c(
      "acute_morbidity",
      "illness_disability",
      "mental_health",
      "years_life_lost"
    ),
    c(
      "Acute morbidity",
      "Comparative illness and disability ratio",
      "Mental health",
      "Years of potential life lost"
    ),
    Weight = c(0.222, 0.294, 0.244, 0.240)
  ),
  aggregate_code = "health",
  aggregate_name = "Health Deprivation and Disability",
  normaliser = "imd_rank_normal"
)
```

The weights add to one, so the weighted sum is also a weighted mean here.

### Crime

Crime follows the same pattern: transform eight indicator ranks to normal scores, then combine them with the published factor-analysis weights.

```{r}
#| label: crime-domain

crime_coin <- make_coin(
  data.frame(
    uCode,
    violence_injury = crime$`Violence with injury, rate per 1,000 'at-risk population'`,
    violence_no_injury = crime$`Violence without injury, rate per 1,000 'at-risk population'`,
    stalking_harassment = crime$`Stalking and harassment, rate per 1,000 'at-risk population'`,
    burglary = crime$`Burglary, rate per 1,000 'at-risk properties'`,
    theft = crime$`Theft, rate per 1,000 'at-risk population'`,
    criminal_damage = crime$`Criminal damage, rate per 1,000 'at-risk population'`,
    public_order_weapons = crime$`Public order and Possession of weapons, rate per 1,000 'at-risk population'`,
    antisocial_behaviour = crime$`Anti-Social Behaviour, rate per 1,000 'at-risk population'`
  ),
  meta(
    c(
      "violence_injury",
      "violence_no_injury",
      "stalking_harassment",
      "burglary",
      "theft",
      "criminal_damage",
      "public_order_weapons",
      "antisocial_behaviour"
    ),
    c(
      "Violence with injury",
      "Violence without injury",
      "Stalking and harassment",
      "Burglary",
      "Theft",
      "Criminal damage",
      "Public order and weapons",
      "Anti-social behaviour"
    ),
    Weight = c(
      0.151,
      0.154,
      0.132,
      0.074,
      0.097,
      0.144,
      0.145,
      0.103
    )
  ),
  aggregate_code = "crime",
  aggregate_name = "Crime",
  normaliser = "imd_rank_normal"
)
```

### Barriers to Housing and Services

Barriers is more hierarchical. We first construct two composite indicators:

- overcrowding, from rooms and bedrooms measures; and
- homelessness, from core and statutory homelessness.

Each pair is transformed to normal scores and combined equally.

```{r}
#| label: barriers-composites

overcrowding_coin <- make_coin(
  data.frame(
    uCode,
    rooms = barriers$`Household overcrowding indicator (rooms)`,
    bedrooms = barriers$`Household overcrowding indicator (bedrooms)`
  ),
  meta(
    c("rooms", "bedrooms"),
    c("Overcrowding: rooms", "Overcrowding: bedrooms"),
    Weight = c(0.5, 0.5)
  ),
  "overcrowding",
  "Household overcrowding",
  "imd_rank_normal"
)

homelessness_coin <- make_coin(
  data.frame(
    uCode,
    core_homelessness = barriers$`Core homelessness rate (% of households)`,
    statutory_homelessness = barriers$`Homelessness indicator (rate per 1,000 households)`
  ),
  meta(
    c("core_homelessness", "statutory_homelessness"),
    c("Core homelessness", "Statutory homelessness"),
    Weight = c(0.5, 0.5)
  ),
  "homelessness",
  "Homelessness",
  "imd_rank_normal"
)
```

The Wider Barriers sub-domain combines five equally weighted indicators. File 8's digital-connectivity field is already expressed in the deprivation direction, so its direction remains positive.

```{r}
#| label: wider-barriers

wider_barriers_coin <- make_coin(
  data.frame(
    uCode,
    housing_affordability = barriers$`Housing affordability indicator`,
    digital_connectivity = barriers$`Digital Connectivity (Broadband download and upload speeds)`,
    patient_gp_ratio = barriers$`Patient-to-GP Ratio`,
    overcrowding = overcrowding_coin$score,
    homelessness = homelessness_coin$score
  ),
  meta(
    c(
      "housing_affordability",
      "digital_connectivity",
      "patient_gp_ratio",
      "overcrowding",
      "homelessness"
    ),
    c(
      "Housing affordability",
      "Digital connectivity deprivation",
      "Patient-to-GP ratio",
      "Household overcrowding",
      "Homelessness"
    ),
    Weight = rep(0.2, 5)
  ),
  "wider_barriers",
  "Wider Barriers",
  "imd_rank_normal"
)
```

The Geographical Barriers sub-domain consists only of the Connectivity Score. Higher connectivity means *less* deprivation, so this indicator has direction `-1`. The two sub-domains are exponentially transformed and combined equally.

```{r}
#| label: barriers-domain

barriers_coin <- make_coin(
  data.frame(
    uCode,
    geographical_barriers = barriers$`Connectivity Score`,
    wider_barriers = wider_barriers_coin$score
  ),
  meta(
    c("geographical_barriers", "wider_barriers"),
    c("Geographical Barriers", "Wider Barriers"),
    Direction = c(-1, 1),
    Weight = c(0.5, 0.5)
  ),
  "barriers",
  "Barriers to Housing and Services",
  "imd_rank_exponential"
)
```

### Living Environment

Living Environment contains equally weighted Indoors and Outdoors indicator groups. Indicators are first mapped to normal scores and combined equally within each sub-domain.

```{r}
#| label: living-subdomains

indoors_coin <- make_coin(
  data.frame(
    uCode,
    housing_condition = living$`Housing in poor condition indicator`,
    energy_performance = living$`Housing energy performance deprivation Score`,
    private_outdoor_space = living$`Housing lacking private outdoor space deprivation score`
  ),
  meta(
    c(
      "housing_condition",
      "energy_performance",
      "private_outdoor_space"
    ),
    c(
      "Housing in poor condition",
      "Housing energy performance deprivation",
      "Housing lacking private outdoor space"
    ),
    Weight = rep(1 / 3, 3)
  ),
  "indoors",
  "Indoors Living Environment",
  "imd_rank_normal"
)

outdoors_coin <- make_coin(
  data.frame(
    uCode,
    noise = living$`Noise pollution`,
    road_casualties = living$`Road traffic casualties involving injury to pedestrians and cyclists`,
    air_quality = living$`Air quality indicator`
  ),
  meta(
    c("noise", "road_casualties", "air_quality"),
    c("Noise pollution", "Road casualties", "Air quality"),
    Weight = rep(1 / 3, 3)
  ),
  "outdoors",
  "Outdoors Living Environment",
  "imd_rank_normal"
)
```

The two sub-domain scores are then exponentially transformed. Indoors receives 70% of the domain weight and Outdoors receives 30%, based on time-use evidence.

```{r}
#| label: living-domain

living_coin <- make_coin(
  data.frame(
    uCode,
    indoors = indoors_coin$score,
    outdoors = outdoors_coin$score
  ),
  meta(
    c("indoors", "outdoors"),
    c("Indoors Living Environment", "Outdoors Living Environment"),
    Weight = c(0.7, 0.3)
  ),
  "living_environment",
  "Living Environment",
  "imd_rank_exponential"
)
```

### Income and Employment

Income and Employment are conceptually simpler: each is a proportion of an at-risk population. In production, however, shrinkage estimation is applied to improve reliability for small areas. File 8 supplies the numerators, but not every input needed to reproduce shrinkage; it also suppresses a small number of numerator values.

We therefore take the released domain rates from File 7. These are substitutions, not reconstructed domains.

```{r}
#| label: income-employment-substitution

official_domain_data <- data.frame(
  uCode,
  income = official$`Income Score (rate)`,
  employment = official$`Employment Score (rate)`,
  education = official$`Education, Skills and Training Score`,
  health = official$`Health Deprivation and Disability Score`,
  crime = official$`Crime Score`,
  barriers = official$`Barriers to Housing and Services Score`,
  living_environment = official$`Living Environment Score`
)

reconstructed_domain_data <- official_domain_data
reconstructed_domain_data$education <- education_coin$score
reconstructed_domain_data$health <- health_coin$score
reconstructed_domain_data$crime <- crime_coin$score
reconstructed_domain_data$barriers <- barriers_coin$score
reconstructed_domain_data$living_environment <- living_coin$score
```

At this point `reconstructed_domain_data` contains two published domain scores, one partly reconstructed domain, and four domains reconstructed from File 8.

## Combine the seven domains into the IMD

The official domain weights are:

| Domain | Weight |
|---|---:|
| Income | 22.5% |
| Employment | 22.5% |
| Education, Skills and Training | 13.5% |
| Health Deprivation and Disability | 13.5% |
| Crime | 9.3% |
| Barriers to Housing and Services | 9.3% |
| Living Environment | 9.3% |

We define these weights in COINr metadata, exponentially transform the seven domain scores, and take their weighted sum.

```{r}
#| label: final-imd

domain_codes <- c(
  "income",
  "employment",
  "education",
  "health",
  "crime",
  "barriers",
  "living_environment"
)

domain_names <- c(
  "Income",
  "Employment",
  "Education, Skills and Training",
  "Health Deprivation and Disability",
  "Crime",
  "Barriers to Housing and Services",
  "Living Environment"
)

domain_weights <- c(
  0.225,
  0.225,
  0.135,
  0.135,
  0.093,
  0.093,
  0.093
)

final_meta <- meta(
  domain_codes,
  domain_names,
  Weight = domain_weights
)

reconstructed_imd_coin <- make_coin(
  reconstructed_domain_data,
  final_meta,
  aggregate_code = "imd",
  aggregate_name = "Index of Multiple Deprivation",
  normaliser = "imd_rank_exponential"
)
```

`reconstructed_imd_coin$coin` retains the raw domains, transformed domains, metadata and final aggregate. The numeric IMD score is available as `reconstructed_imd_coin$score`.

## Calculate ranks and deciles

The official rank assigns 1 to the most deprived LSOA. Sorting the input by LSOA code before ranking also gives deterministic ordering if reconstructed scores tie.

```{r}
#| label: ranks-deciles

rank_desc <- function(x) {
  as.integer(rank(-x, ties.method = "first"))
}

rank_to_decile <- function(x, n = length(x)) {
  pmin(10L, as.integer(ceiling(x * 10 / n)))
}

reconstructed_rank <- rank_desc(reconstructed_imd_coin$score)
reconstructed_decile <- rank_to_decile(reconstructed_rank)

reconstructed_results <- tibble(
  lsoa21_code = uCode,
  imd_score = reconstructed_imd_coin$score,
  imd_rank = reconstructed_rank,
  imd_decile = reconstructed_decile
)

head(reconstructed_results)
```

## Validate the reconstruction

### Domain-level agreement

First compare each reconstructed domain with its published score. Income and Employment match by construction because they are substitutions.

```{r}
#| label: domain-validation

official_domain_matrix <- official_domain_data[-1]
reconstructed_domain_matrix <- reconstructed_domain_data[-1]

domain_validation <- bind_rows(lapply(seq_along(domain_codes), function(j) {
  reconstructed_score <- reconstructed_domain_matrix[[j]]
  official_score <- official_domain_matrix[[j]]

  tibble(
    domain = domain_names[j],
    rank_spearman = cor(
      reconstructed_score,
      official_score,
      method = "spearman"
    ),
    mean_absolute_score_difference = mean(
      abs(reconstructed_score - official_score)
    )
  )
}))

knitr::kable(domain_validation, digits = 6)
```

Residual differences reflect rounding of released indicators and missing intermediate pre-shrinkage values. The correlations remain very close to one.

### Final score, rank and decile agreement

```{r}
#| label: final-validation

official_imd_score <-
  official$`Index of Multiple Deprivation (IMD) Score`
official_imd_rank <-
  official$`Index of Multiple Deprivation (IMD) Rank (where 1 is most deprived)`
official_imd_decile <-
  official$`Index of Multiple Deprivation (IMD) Decile (where 1 is most deprived 10% of LSOAs)`

validation <- tibble(
  score_pearson = cor(
    reconstructed_imd_coin$score,
    official_imd_score
  ),
  rank_spearman = cor(
    reconstructed_rank,
    official_imd_rank,
    method = "spearman"
  ),
  mean_absolute_rank_difference = mean(
    abs(reconstructed_rank - official_imd_rank)
  ),
  maximum_absolute_rank_difference = max(
    abs(reconstructed_rank - official_imd_rank)
  ),
  exact_rank_match = mean(
    reconstructed_rank == official_imd_rank
  ),
  decile_match = mean(
    reconstructed_decile == official_imd_decile
  )
)

validation |>
  mutate(
    exact_rank_match = 100 * exact_rank_match,
    decile_match = 100 * decile_match
  ) |>
  rename(
    `Exact rank match (%)` = exact_rank_match,
    `Decile match (%)` = decile_match
  ) |>
  knitr::kable(digits = 4)
```

The reconstructed scores have a Pearson correlation above 0.999999 with the published IMD. More than 99.6% of LSOAs are assigned to the same decile. Exact rank agreement is much lower because small score differences can reorder near-tied LSOAs without changing their practical deprivation classification.

```{r}
#| label: fig-rank-comparison
#| fig-cap: "Published and reconstructed IMD ranks. Points lying on the diagonal have identical ranks."
#| fig-alt: "Scatter plot showing reconstructed ranks closely following published ranks along a diagonal line."
#| fig-width: 7
#| fig-height: 6

plot(
  official_imd_rank,
  reconstructed_rank,
  pch = 16,
  cex = 0.25,
  col = rgb(0.1, 0.3, 0.6, 0.15),
  xlab = "Published IMD rank",
  ylab = "Reconstructed IMD rank"
)
abline(0, 1, col = "#D55E00", lwd = 2)
```

## Isolate the effect of published rounding

File 9 contains the seven exponentially transformed domain scores. Starting from these removes nearly all discrepancies arising earlier in the pipeline and tests only the final weighting step.

```{r}
#| label: transformed-domain-benchmark

transformed <- read_excel(
  source_paths[["transformed_domains"]],
  sheet = "IoD25 Transformed Domain Scores"
) |>
  transmute(
    uCode = `LSOA code (2021)`,
    income = `Income Score - exponentially transformed`,
    employment = `Employment Score - exponentially transformed`,
    education = `Education Score - exponentially transformed`,
    health = `Health Score - exponentially transformed`,
    crime = `Crime Score - exponentially transformed`,
    barriers = `Barriers Score - exponentially transformed`,
    living_environment = `Living Environment Score - exponentially transformed`
  ) |>
  arrange(uCode)

stopifnot(identical(transformed$uCode, uCode))

transformed_benchmark_coin <- make_coin(
  transformed,
  final_meta,
  aggregate_code = "imd",
  aggregate_name = "Index of Multiple Deprivation"
)

benchmark_rank <- rank_desc(transformed_benchmark_coin$score)

benchmark_validation <- tibble(
  maximum_absolute_score_difference = max(abs(
    transformed_benchmark_coin$score - official_imd_score
  )),
  exact_rank_match = mean(benchmark_rank == official_imd_rank),
  mean_absolute_rank_difference = mean(abs(
    benchmark_rank - official_imd_rank
  )),
  maximum_absolute_rank_difference = max(abs(
    benchmark_rank - official_imd_rank
  ))
)

benchmark_validation |>
  mutate(exact_rank_match = 100 * exact_rank_match) |>
  rename(`Exact rank match (%)` = exact_rank_match) |>
  knitr::kable(digits = 4)
```

This final-stage benchmark reproduces about 90.4% of ranks exactly and differs by at most three rank places. Its maximum score difference is less than 0.001. That residual is consistent with the precision of the released transformed-domain workbook.

## What the reconstruction demonstrates

The exercise separates three ideas that are easy to conflate:

- **Reconstruction of the method.** COINr reproduces the IMD hierarchy, indicator directions, rank-normal transformations, exponential transformations and weights.
- **Reconstruction from public inputs.** Health, Crime, Barriers and Living Environment can be rebuilt extremely closely from released indicators. Education can be partly rebuilt.
- **Independent replication.** This is not possible for Income, Employment and the Children and Young People education sub-domain because essential inputs are not public.

The distinction matters. A high correlation does not turn a published-score substitution into an independently reconstructed component.

## Package integration {#package-integration}

To add this walkthrough to `{IMD}` as a vignette:

1. copy this file to `vignettes/replicate-imd2025-with-coinr.qmd`;
2. add `COINr`, `dplyr`, `readxl`, `knitr` and `quarto` to `Suggests` in `DESCRIPTION`, and set `VignetteBuilder: quarto`;
3. make the source data available offline, preferably as cleaned package data objects;
4. replace the download chunk with those package objects; and
5. render the vignette during development with:

```sh
quarto render vignettes/replicate-imd2025-with-coinr.qmd
```

The current `imd2025_england_lsoa21_indicators` package object contains the Health, Crime and Barriers inputs used above, but does not contain the Living Environment indicators or the published Children and Young People sub-domain score. Those fields need to be added as package data, or supplied under `inst/extdata`, before removing the source-file ingestion code.

For a lightweight package vignette, consider pre-computing the three validation tables as package data while leaving all reconstruction code visible with `eval: false`. For an executable vignette, keep the embedded checks on LSOA count, code alignment and validation tolerances so upstream source changes fail loudly.

## Session information

```{r}
#| label: session-info

sessionInfo()
```
